---
title: "Applying the Eigendecomposition of a Matrix to Generate Fibonacci Numbers"
author: "by Matt Cook"
output:
  html_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

<br> 

## Fibonacci Numbers

Fibonacci Numbers are numbers generated from a special sequence. The sequence begins 0, 1, 1, 2, 3, 5, 8, 13, 21, 34 ... and continues forever.

The sequence starts with 0 and 1, and each subsequent number is produced by summing the two previous numbers.

Formally,

$$x_i=x_{i-1} + x_{i-2}$$

where *x~i~* is the *i^th^* Fibonacci number, and *x~i-1~* and *x~i-2~* are the two previous numbers in the sequence.

<br>

It is simple to compute the Fibonacci numbers by hand (or to write a computer program to do it). However, to compute a given Fibonacci number, these methods require computing all the previous numbers in the sequence.

I will present a well-known method of using linear algebra for computing any given Fibonacci number in the sequence automatiucally, without having to generate the sequence leading up to that number.

In other words, can we, in a single step, compute the 50^th^ Fibonacci Number?

<br>

## Applying Linear Algebra to Generate Fibonacci Numbers

The following matrix **A** is useful to us: 

$$A = \begin{bmatrix}
0 & 1\\
1 & 1
\end{bmatrix}$$

If we represent the first two Fibonacci numbers (e.g., 0 & 1) as a vector and multiply it with our matrix **A** two useful things happen:

$$\begin{bmatrix}
0 & 1\\
1 & 1
\end{bmatrix}\begin{bmatrix}
0\\
1
\end{bmatrix}=\begin{bmatrix}
1\\
1
\end{bmatrix}$$

First, the sum of these two numbers 0 + 1 = 1 appears as the second element in our new vector. This is useful because this is the next number in our sequence.

Second, a 1 also appears as our first element in our new vector. This is useful because our vector now has 1 and 1 (the second and third Fibonacci numbers), which allows us to repeat the process to produce the next number in the sequence:

$$\begin{bmatrix}
0 & 1\\
1 & 1
\end{bmatrix}\begin{bmatrix}
1\\
1
\end{bmatrix}=\begin{bmatrix}
1\\
2
\end{bmatrix}$$


Which we can repeat a third time:

$$\begin{bmatrix}
0 & 1\\
1 & 1
\end{bmatrix}\begin{bmatrix}
1\\
2
\end{bmatrix}=\begin{bmatrix}
2\\
3
\end{bmatrix}$$

<br>

To do this all in one step, we can multiply our original vector of the first two Fibonacci numbers (i.e., 0 & 1) by our matrix **A** three times:

$$\begin{bmatrix}
0 & 1\\
1 & 1
\end{bmatrix}\begin{bmatrix}
0 & 1\\
1 & 1
\end{bmatrix}\begin{bmatrix}
0 & 1\\
1 & 1
\end{bmatrix}\begin{bmatrix}
1\\
2
\end{bmatrix}=\begin{bmatrix}
2\\
3
\end{bmatrix}$$

This allows us to compute the fifth Fibonacci number. We can write things more compactly:

$$\begin{bmatrix}
0 & 1\\
1 & 1
\end{bmatrix}^3\begin{bmatrix}
1\\
1
\end{bmatrix}=\begin{bmatrix}
2\\
3
\end{bmatrix}$$

Here we compute the powers of a matrix. By computing the third power of the matrix **A**, we can compute the fifth Fibonacci number.

<br>

Similarily, computing the 50^th^ power of the matrix **A**, we can compute the 52^nd^ Fibonacci number (as well as the 51^st^):

$$\begin{bmatrix}
0 & 1\\
1 & 1
\end{bmatrix}^{50}\begin{bmatrix}
1\\
1
\end{bmatrix}=\begin{bmatrix}
12586269025\\
20365011074
\end{bmatrix}$$

<br>

## Using the Eigendecomposition of a Matrix to Compute Matrix Powers

The Eigendecomposition of a square matrix decomposes a matrix into three more fundamental matrices. When these three matrices are multiplied together they produce the original matrix. This is similar to how a composite number like 30 can be decomposed into three more fundamental prime numbers (30 = 2 * 3 * 5). 

$$A = Q \Lambda Q^{-1}$$

In the case of our matrix **A**, the following equation evaluates to:

$$\begin{bmatrix}
0 & 1\\
1 & 1
\end{bmatrix}=\begin{bmatrix}
0.5257311 & -0.8506508\\
0.8506508 & 0.5257311
\end{bmatrix}\begin{bmatrix}
1.618034 & 0\\
0 & -0.618034
\end{bmatrix}\begin{bmatrix}
0.5257311 & 0.8506508\\
-0.8506508 & 0.5257311
\end{bmatrix}$$

The columns in the first matrix, **Q** are the *eigenvectors* of the matrix **A**. The second matrix $\Lambda$, is a diagonal matrix which stores the *eigenvalues* of **A**. The third matrix **$Q^{-1}$** is the *inverse* of **Q**. 

That is, A is re-represented as a rotation (**Q**), a scaling ($\Lambda$), and another rotation (**$Q^{-1}$**).

This is also useful to us. We can now compute any power of the matrix **A** by simply computing the powers of the eigenvalues in $\Lambda$.

As an example, we can compute the 50th power of **A** with the following:

$$\begin{bmatrix}
0 & 1\\
1 & 1
\end{bmatrix}^{50}=\begin{bmatrix}
0.5257311 & -0.8506508\\
0.8506508 & 0.5257311
\end{bmatrix}\begin{bmatrix}
1.618034 & 0\\
0 & -0.618034
\end{bmatrix}^{50}\begin{bmatrix}
0.5257311 & 0.8506508\\
-0.8506508 & 0.5257311
\end{bmatrix}$$

Importantly, the powers of the diagonal matrix $\Lambda$ are easily computed by simply computing the powers of the number in the diagonals (e.g., 1.618034^50^ and -0.618034^50^)

<br>

## Applying the Eigendecomposition to Generate any Fibonacci Number

We can now combine a few of these ideas to very simply compute any Fibonacci number without needing to compute all the numbers of the sequence that came before.

The following will compute any Fibonacci number automatically!

$$\begin{bmatrix}
fib^{n+1}\\
fib^{n+2}
\end{bmatrix}=\begin{bmatrix}
0.5257311 & -0.8506508\\
0.8506508 & 0.5257311
\end{bmatrix}\begin{bmatrix}
1.618034 & 0\\
0 & -0.618034
\end{bmatrix}^{n}\begin{bmatrix}
0.5257311 & 0.8506508\\
-0.8506508 & 0.5257311
\end{bmatrix}\begin{bmatrix}
0\\
1
\end{bmatrix}$$

As an example, the following will compute the 51^st^ and 52^st^ Fibonacci numbers that we computed before with only a handful of matrix operations (compared to the 50+ matrix operations without using the eigendecomposition of the matrix **A**):

$$\begin{bmatrix}
fib^{51}\\
fib^{52}
\end{bmatrix}=\begin{bmatrix}
0.5257311 & -0.8506508\\
0.8506508 & 0.5257311
\end{bmatrix}\begin{bmatrix}
1.618034 & 0\\
0 & -0.618034
\end{bmatrix}^{50}\begin{bmatrix}
0.5257311 & 0.8506508\\
-0.8506508 & 0.5257311
\end{bmatrix}\begin{bmatrix}
0\\
1
\end{bmatrix}=\begin{bmatrix}
12586269025\\
20365011074
\end{bmatrix}$$

<br>

## Importance

Matrices are used to represent data applicable in science, engineering, data-science, and machine learning. By re-representing the data (i.e., as the eigen vectors and values of a matrix) we can more easily understand and process the data. This re-representation allows us to solve problems that would be difficult, time-consuming, or impossible to solve using other methods.

<br>

## R Code

```{r}
# 2x2 matrix
a <- matrix(c(0, 1,
              1, 1), 
            2, 2)

#compute eigen vectors and values
eig <- eigen(a)

#compute the n+1th and n+2th Fibonacci numbers
eig$vectors %*% diag(eig$values^50) %*% solve(eig$vectors) %*% c(0, 1)

```



<br>
<br>
<br>
<br>
<br>